Beliefs about the development of mental life

### 2021-10-16 paper/paper.Rmd
revisions for submission

---
title: "Beliefs about the development of mental life"
author:
  - Kara Weisman^[Department of Psychology, University of California, Riverside; kgweisman@gmail.com]
  - Lucy S. King^[Department of Psychiatry and Behavioral Sciences, Tulane University School of Medicine; Department of Psychology, The University of Texas at Austin]
  - Kathryn L. Humphreys^[Department of Psychology, Vanderbilt University]
output:
  # word_document:
  #   reference_docx: word-styles-ref.docx
  bookdown::pdf_document2:
    latex_engine: xelatex
    fig_caption: yes
    toc: false
    number_sections: false
bibliography: "../supplement/references.bib"
cls: "../supplement/article.cls"
figsintext: yes
header-includes: 
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{float}
---

@@

In a series of large-scale studies we assessed how US adults conceptualize the development of the human mind over the first five years of life. Exploratory factor analysis identified four categories of mental capacities that anchored participants’ representations of the developing human mind: _bodily sensations_ (e.g., hunger, pain), _negative affect_ (e.g., distress, frustration), _social connection_ (e.g., love, learning from others), and _cognition and control_ (e.g., planning, self-control). Participants perceived that these four aspects of mental life were present to different degrees at birth, followed different developmental trajectories, and were driven by different developmental mechanisms (such as biological "preprogramming," physical maturation, passive observation, and social learning). These studies reveal the intuitive theories that govern US adults’ understanding of the development of the human mind, illuminating the cognitive architecture that supports some of the most important social interactions: caregiving relationships with infants and children.

## Significance statement

Caregiving relationships with infants and children are among the most common, important, and complex human social interactions. Caring for a young child over time requires an adult to continually update their own mental representations of the child's capacities as the child grows and changes. These representations are likely to have important consequences for the child's well-being—particularly in the first few years of life, when the child's communication skills are limited and the caregiver must make inferences about the child's thoughts, feelings, and needs. The current studies shed light on this fascinating and understudied aspect of "mind perception," in turn laying the foundation for theory-based interventions to encourage positive and responsive parenting behaviors.

\newpage

Frustration, compassion, worry, humor, imagination, love—such experiences are fundamental parts of human life. But few people would assert that these capacities are fully developed at birth. In any given interaction with a child, adults come face to face with a being whose experience of the world is like their own, but different; whose mind is capable of some of the thoughts and feelings that an adult’s mind is, but perhaps not all of them. When an adult interacts with the same child over time, they witness innumerable changes in that child’s mental life: A sleepy newborn gradually becomes a child who experiences complex emotions and original ideas and develops personal memories and intricate plans. This familiar but remarkable transformation requires adults to make inferences about a given child’s developing mental capacities in order to interpret that child’s behaviors and respond appropriately.

In this paper, we examine how adults conceptualize the developing minds of infants and children. We draw on techniques developed in studies of mind perception, in which researchers use large datasets and modern statistical methods to reveal conceptual representations of "the mind," broadly construed. In a groundbreaking study, Gray et al. asked participants to compare the mental capacities of a range of humans, animals, technologies, and other entities, and argued that participants’ judgments revealed a fundamental distinction between "agency" (e.g., capacities for self-control, moral reasoning, memory, and planning) and "experience" (e.g., capacities for hunger, fear, pain, and pleasure) [@gray.etal2007]. More recently, Weisman et al. used a similar empirical approach to argue that people’s conceptions of mental life are anchored by three fundamental components: "body" (physiological sensations such as hunger and pain), "heart" (social-emotional abilities such as embarrassment and pride), and "mind" (perceptual-cognitive capacities such as vision and memory [@weisman.etal2017a]; see [@malle2019] for a similar three-factor solution. Such findings have inspired subsequent work on topics as wide-ranging as moral reasoning, dehumanization and stereotyping, beliefs about the afterlife, and human-robot interactions.

However, the burgeoning field of mind perception has largely neglected the question of how the human mind emerges over development. In many studies, infants and children have simply not been included as targets for mental capacity attributions; instead, participants have been asked to reason about human adults alongside a range of non-human entities, including other animals, technologies, and supernatural beings. Even studies that have probed participants’ mental capacity attributions to infants and children have generally treated them as beings distinct from adults, rather than as snapshots of a single mind as it emerges and changes over time. For example, Gray et al.’s early work on mind perception suggested that a five-month-old infant is an entity that is generally considered to have experience but not agency—in contrast to, say, a five-year-old child (considered to have experience as well as limited agency), a human adult (considered to have maximal experience and agency), or a robot (considered to have agency but not experience) [@gray.etal2007]. Likewise, Weisman et al.’s studies characterized infants and young children—along with, e.g., chimpanzees, elephants, and dolphins—as entities that are perceived to be fully capable of physiological sensations but more limited in their perceptual-cognitive capacities and social-emotional abilities; this is in contrast to human adults, who are perceived to be fully capable across these domains [@weisman.etal2017a]. Although such findings can be reinterpreted as providing preliminary evidence that people consider the minds of infants and children to differ from the minds of adults, they do not address how people conceptualize the development of the mind in early life.

Yet infants and young children are far from unusual social partners. Caregiving relationships are at the core of human existence; for many people, infants and young children are among the most frequent and highly valued minds that we encounter. Caring for a young child involves making countless decisions under uncertainty: Why is my child crying, smiling, shouting, or biting? Can they understand what I am saying? Does my child love me? As in other domains of reasoning [@gerstenberg.tenenbaum2017; @gopnik.wellman1994; @wellman.gelman1992], people likely draw on folk beliefs and intuitive theories about how the mind develops to predict, explain, and respond to children’s behavior and development, with consequences for the well-being of children in their care.

In the current studies we set out to investigate this important and understudied aspect of mind perception: how ordinary people reason about the development of human mental life. In three studies, we assessed US adults’ folk beliefs about children’s developing abilities for physiological sensation, perception, cognition, emotion, self-regulation, social interaction, and other aspects of "mental life," broadly construed. We focused on development between birth and age five years—a time of particularly dramatic maturation in perceptual acuity, emotional experience, cognitive ability, executive function, social engagement, and many other domains.

# Results

## Study 1

We began with an expansive exploration of US adults’ attributions of mental life to infants and young children of different ages. Drawing inspiration from previous work on mind perception [@gray.etal2007; @weisman.etal2017a; @weisman.etal2017b; @weisman.etal2018] as well as standard assessments of infant temperament and behavior [@putnam.etal2006; @rothbart1978], we included a wide range of 60 capacities in this initial exploration, from basic physiological sensations, to capacities for perception, cognition, and emotion, to abilities for self-regulation and social interaction. We asked participants ($n$ = `r s1_results$demo %>% distinct() %>% nrow()`) to assess these capacities at three different ages (at birth, 9 months, and 5 years), with the goal of characterizing the conceptual structure underlying participants’ reasoning about the development of mental life: What do people perceive to be the fundamental components of mental life, as they pertain to a developing human child?

We designed Study 1 to explore the correlational structure of participants’ responses, as a window into the underlying conceptual structure that might support their reasoning about the development of mental life. Our primary analysis for this preregistered study was an exploratory factor analysis (EFA) of participants’ capacity ratings. This analysis allowed us to examine which capacities tend to "hang together" in participants’ assessments of the mental lives of infants and young children. For example, when a participant indicated that newborns are highly capable of feeling scared, what other capacities did they tend to attribute to newborns? Following previous work on mind perception, we argue that the suites of capacities revealed by applying dimensionality reduction (in our case, EFA) to this covariance structure offer a meaningful approximation of the latent conceptual structure underlying participants’ reasoning about the developing human mind.

A `r s1_results$efa$factors`-factor solution (as suggested by parallel analysis) accounted for `r round_x(s1_results$efa$Vaccounted["Cumulative Var", "MR4"]*100, 0)`% of the total variance in participants’ capacity attributions. See Figure \@ref(fig:fig01) for all factor loadings after oblique transformation.

The first factor corresponded primarily to capacities related to thinking, reasoning, agency, and executive function—a suite of capacities that we will refer to as _cognition and control_. It was the dominant factor for such items as "planning," "having self-control," "thinking before they act," "having goals," "reasoning about things," "controlling their emotions," "telling right from wrong," and "understanding what somebody else is thinking," and accounted for `r round_x(s1_results$efa$Vaccounted["Proportion Explained", "MR1"]*100, 0)`% of the shared variance in the rotated `r s1_results$efa$factors`-factor solution and `r round_x(s1_results$efa$Vaccounted["Proportion Var", "MR1"]*100, 0)`% of the total variance in participants’ capacity attributions.

The second factor corresponded primarily to social abilities and emotional experiences, particularly those with positive valence—a suite of capacities that we will refer to as _social connection_. It was the dominant factor for such items as "feeling excited", "finding something funny", "loving somebody", "learning from other people", "feeling happy", "feeling loved", "recognizing somebody else", and "getting pleasure from music", and accounted for `r round_x(s1_results$efa$Vaccounted["Proportion Explained", "MR3"]*100, 0)`% of the shared variance in the rotated `r s1_results$efa$factors`-factor solution and `r round_x(s1_results$efa$Vaccounted["Proportion Var", "MR3"]*100, 0)`% of the total variance in participants’ capacity attributions.

The third factor corresponded primarily to physiological sensations related to biological needs—a suite of capacities that we will refer to as _bodily sensations_. It was the dominant factor for such items as "getting hungry", "feeling pain", "feeling tired", "feeling thirsty", "feeling too hot or too cold", "feeling physically uncomfortable", "hearing sounds", and "being comforted by physical touch", and accounted for `r round_x(s1_results$efa$Vaccounted["Proportion Explained", "MR2"]*100, 0)`% of the shared variance in the rotated `r s1_results$efa$factors`-factor solution and `r round_x(s1_results$efa$Vaccounted["Proportion Var", "MR2"]*100, 0)`% of the total variance in participants’ capacity attributions.

Finally, the fourth factor corresponded to negatively-valenced sensations and emotions—a suite of capacities that we will refer to as _negative affect_. It was the dominant factor for the items "feeling helpless", "feeling overwhelmed", "feeling frustrated", "feeling annoyed", and "feeling neglected"; the items "feeling distressed", "feeling confused", "feeling worried", and "feeling lonely" also loaded relatively strongly on this factor (albeit slightly more strongly on other factors). This factor accounted for `r round_x(s1_results$efa$Vaccounted["Proportion Explained", "MR4"]*100, 0)`% of the shared variance in the rotated `r s1_results$efa$factors`-factor solution and `r round_x(s1_results$efa$Vaccounted["Proportion Var", "MR4"]*100, 0)`% of the total variance in participants’ capacity attributions.

How robust are these findings? The factors we have called _bodily sensations_, _social connection_, and _cognition and control_ each had strong factor loadings ($\geq$ 0.60) for many individual items ($n$ = `r min(s1_strong_loadings$n)`-`r max(s1_strong_loadings$n)`) and each accounted for a substantial amount of the total variance in participants’ capacity attributions (`r round_x(min(s1_results$efa$Vaccounted["Proportion Var",c("MR1","MR3", "MR2")])*100, 0)`-`r round_x(max(s1_results$efa$Vaccounted["Proportion Var",c("MR1","MR3", "MR2")])*100, 0)`%), indicating three robustly distinct and interpretable factors. In contrast, the fourth factor—_negative affect_—had more moderate loadings even for its strongest-loading items (all loadings < `r abs(s1_results$efa$loadings[][,"MR4"]) %>% max() %>% round(., 2) + 0.01`) and accounted for only `r round_x(s1_results$efa$Vaccounted["Proportion Var","MR4"]*100, 0)`% of the total variance. Nonetheless, we consider the general domain of _negative affect_ to be of particular theoretical and clinical interest.

What does it mean that these four factors emerged? At a most basic level, our results reflect sets of intercorrelated capacities: When a participant judged that a child of a certain age was highly capable of planning, they also tended to judge that that child was capable of having self control, thinking before they act, and the other capacities with strong factor loadings on the _cognition and control_ factor; and likewise for the suites of capacities that define the other factors revealed by EFA. Therefore, we take these factors to reflect latent constructs—fundamental components of mental life relevant to reasoning about human children. Moreover, because each participant was asked to reason not just about some class of target characters (e.g., "chimpanzees," "robots"), but about the human child over development, we further propose that part of what contributed to the differentiation of these four factors was participants’ perceptions of development: the extent to which different capacities are present at birth, the rate at which different capacities develop, and the mechanisms that drive this development. In Studies 2 and 3, we explore these possibilities in greater depth.

## Study 2

In Study 2 ($n$ = `r s2_results$demo %>% distinct() %>% nrow()`), we replicated the conceptual structure identified by Study 1 and then used it to chart how different aspects of mental life are perceived to change over development: What kinds of abilities do people believe are present at birth, and to what degree? To what extent are capacities for _bodily sensations_, _negative affect_, _social connection_, and _cognition and control_ perceived to change over childhood, and what is the shape of this perceived developmental trajectory?

A preregistered EFA yielded four factors very similar to the four factors from Study 1, with all capacity items loading strongly on the factors that they were selected to represent and less strongly on other factors (see Supplemental Materials). Minimally, this suggests that our selection of mental capacities for this study captured the essential meaning of the four factors revealed in Study 1. Beyond this, we consider these results to be a somewhat independent conceptual replication of Study 1 results and validation of our interpretation of these factors. After all, given that Study 2 featured only 20 of the 60 capacities featured in Study 1, it certainly could have been the case that this analysis would suggest retaining fewer factors (as often occurs with fewer variables). Conversely, given that Study 2 featured 13 within-subjects observations for each capacity compared to the 3 in Study 1, this analysis could have surfaced more than four factors (as often occurs with more observations). Finally, if we had misinterpreted the fundamental semantic content that distinguished the factors in Study 1 and selected inappropriate capacities to represent these factors, the Study 2 analysis could have identified more capacities with cross-loadings. Instead, the results of this analysis offer some validation of the four factors we have called _bodily sensations_, _negative affect_, _social connection_, and _cognition and control_.

Regressing participants' item-level responses onto target age, domain, and interactions between them via a multilevel generalized additive model confirmed both of our preregistered hypotheses: On average, participants rated older children as more capable, but perceptions of the development of children’s capacities differed dramatically across domains. On one extreme, participants attributed some, but relatively little, capacity for _cognition and control_ to newborns (median score for newborns: `r s2_cap_rat_newborns_ss %>% filter(factor == "Cognition and control") %>% pull(median) %>% round_x(., 2)` out of 100 points, $M$ = `r s2_cap_rat_newborns_ss %>% filter(factor == "Cognition and control") %>% pull(mean) %>% round_x(., 2)`, $sd$ = `r s2_cap_rat_newborns_ss %>% filter(factor == "Cognition and control") %>% pull(sd) %>% round_x(., 2)`), but reported that such capacities increase dramatically over the target age range (average smoothing term, analogous to a slope in linear regression: $F$ = `r round_x(s2_gam_stab["s(target_year):factorCognition and control", "F"], 2)`, $p$ `r ifelse(s2_gam_stab["s(target_year):factorCognition and control", "p-value"] < 0.001, "< 0.001", paste("=", round_x(s2_gam_stab["s(target_year):factorCognition and control", "p"], 3)))`, $edf$ = `r round_x(s2_gam_stab["s(target_year):factorCognition and control", "edf"], 2)`). Compared to these attributions of _cognition and control_, participants attributed substantially greater capacities to newborns in the domains of _social connection_ (median score for newborns: `r s2_cap_rat_newborns_ss %>% filter(factor == "Social connection") %>% pull(median) %>% round_x(., 2)` out of 100 points, $M$ = `r s2_cap_rat_newborns_ss %>% filter(factor == "Social connection") %>% pull(mean) %>% round_x(., 2)`, $sd$ = `r s2_cap_rat_newborns_ss %>% filter(factor == "Social connection") %>% pull(sd) %>% round_x(., 2)`; increase in log-odds relative to _cognition and control_: $\beta$ = `r round_x(s2_gam_ptab["factorsoc_cog", "Estimate"], 2)`, $p$ `r ifelse(s2_gam_ptab["factorsoc_cog", "Pr(>|t|)"] < 0.001, "< 0.001", paste("=", round_x(s2_gam_ptab["factorsoc_cog", "Pr(>|t|)"], 3)))`) and _negative affect_ (median score for newborns: `r s2_cap_rat_newborns_ss %>% filter(factor == "Negative affect") %>% pull(median) %>% round_x(., 2)` out of 100 points, $M$ = `r s2_cap_rat_newborns_ss %>% filter(factor == "Negative affect") %>% pull(mean) %>% round_x(., 2)`, $sd$ = `r s2_cap_rat_newborns_ss %>% filter(factor == "Negative affect") %>% pull(sd) %>% round_x(., 2)`; increase in log-odds: $\beta$ = `r round_x(s2_gam_ptab["factorneg_cog", "Estimate"], 2)`, $p$ `r ifelse(s2_gam_ptab["factorneg_cog", "Pr(>|t|)"] < 0.001, "< 0.001", paste("=", round_x(s2_gam_ptab["factorneg_cog", "Pr(>|t|)"], 3)))`) and, accordingly, reported relatively small increases over the target age range in both domains (average smoothing terms for _social connection_: $F$ = `r round_x(s2_gam_stab["s(target_year):factorSocial connection", "F"], 2)`, $p$ `r ifelse(s2_gam_stab["s(target_year):factorSocial connection", "p-value"] < 0.001, "< 0.001", paste("=",  round_x(s2_gam_stab["s(target_year):factorCognition and control", "p"], 3)))`, $edf$ = `r round_x(s2_gam_stab["s(target_year):factorSocial connection", "edf"], 2)`; for _negative affect_: $F$ = `r round_x(s2_gam_stab["s(target_year):factorNegative affect", "F"], 2)`, $p$ `r ifelse(s2_gam_stab["s(target_year):factorNegative affect", "p-value"] < 0.001, "< 0.001", paste("=", round_x(s2_gam_stab["s(target_year):factorNegative affect", "p"], 3)))`, $edf$ = `r round_x(s2_gam_stab["s(target_year):factorCognition and control", "edf"], 2)`). At the other extreme, in the domain of _bodily sensations_ participants attributed nearly maximal capacity to newborns (median score for newborns: `r s2_cap_rat_newborns_ss %>% filter(factor == "Bodily sensations") %>% pull(median) %>% round_x(., 2)` out of 100 points, $M$ = `r s2_cap_rat_newborns_ss %>% filter(factor == "Bodily sensations") %>% pull(mean) %>% round_x(., 2)`, $sd$ = `r s2_cap_rat_newborns_ss %>% filter(factor == "Bodily sensations") %>% pull(sd) %>% round_x(., 2)`; increase in log-odds relative to _cognition and control_: $\beta$ = `r round_x(s2_gam_ptab["factorbod_cog", "Estimate"], 2)`, $p$ `r ifelse(s2_gam_ptab["factorbod_cog", "Pr(>|t|)"] < 0.001, "< 0.001", paste("=", round_x(s2_gam_ptab["factorbod_cog", "Pr(>|t|)"], 3)))`) and a significant but small increase in capacities for _bodily sensations_ over the target age range (average smoothing term: $F$ = `r round_x(s2_gam_stab["s(target_year):factorBodily sensations", "F"], 2)`, $p$ `r ifelse(s2_gam_stab["s(target_year):factorBodily sensations", "p-value"] < 0.001, "< 0.001", paste("=", round_x(s2_gam_stab["s(target_year):factorBodily sensations", "p"], 3)))`, $edf$ = `r round_x(s2_gam_stab["s(target_year):factorBodily sensations", "edf"], 2)`).

Visual inspection of Figure \@ref(fig:fig02), Panel A, clearly illustrates that the shape of perceived growth in children’s capacities varied substantially across domains. In the domain of _bodily sensations_, participants perceived hardly any growth over the target age range, and what growth they did perceive occurred in the first year of life. Participants perceived development in the domains of _negative affect_ and _social connection_ to be quite non-linear, characterized by rapid development over the first 1-2 years of life, at which point the curves level off dramatically. In contrast, participants perceived dramatic development in the domain of _cognition and control_, but perceived this development to unfold fairly steadily across the full age range.

We speculate that these differences in perceptions of growth in children’s capacities emerge from intuitive theories of development that specify different mechanisms or drivers of development for different domains. For example, if participants believe that development in the domain of _bodily sensations_ is primarily driven by innate biological forces, they might assume that most of this development is complete before birth, leaving little room for change over the target ages assessed (see @berent.etal2019 for empirical evidence in line with this speculation). In contrast, in a domain where participants believe development depends on observation, exploration, or social learning, they might think that children need time in the world to change and grow, yielding lower estimates of capacities at birth and slower estimates of rates of change; or, in a domain where participants believe that development occurs through explicitly teaching the child something, they might perceive that development proceeds nonlinearly, such that large changes occur following specific milestones (e.g., entry to preschool). We explore this possibility in Study 3.

## Study 3

Study 3 was designed with two goals in mind: First, we aimed to provide a more direct assessment of the differences in perceived developmental trajectories surfaced by Study 2. Second, we aimed to probe the intuitive theories that might underlie the differentiation of the four factors surfaced by Studies 1-2 and the perceived developmental trajectories surfaced by Study 2 (and the current study).

In Study 2, participants focused on one target age at a time, assessing a variety of capacities for that age before proceeding to the next target age (from youngest to oldest). This design was intended to encourage participants to focus primarily on providing a holistic assessment of mental life at a given target age; the trajectories connecting target ages were likely not the participants’ primary concern in the moment of responding. In Study 3, we flipped the design to focus participants’ attention on development more explicitly: Participants were asked to provide judgments of a single capacity for the full range of target ages (arranged from youngest to oldest) on the same "trial," before proceeding to the next capacity. In addition to signaling more strongly to participants that we were interested in their perceptions of development, this approach gave participants the opportunity to view and adjust all of their responses for a given target capacity simultaneously, charting out their perceptions of the full developmental trajectory from birth to age 5 years. In order to accommodate the addition of questions probing intuitive theories of development (see next section), participants assessed only 2 capacities per domain; see Methods.

This approach yielded a very similar picture of US adults’ understanding of the development of human mental life; see Figure \@ref(fig:fig02), Panel B. As in Study 2, participants perceived that newborns have very little capacity for _cognition and control_ (median score for newborns: `r s3_cap_rat_newborns_ss %>% filter(factor == "Cognition and control") %>% pull(median) %>% round_x(., 2)` out of 100 points, $M$ = `r s3_cap_rat_newborns_ss %>% filter(factor == "Cognition and control") %>% pull(mean) %>% round_x(., 2)`, $sd$ = `r s3_cap_rat_newborns_ss %>% filter(factor == "Cognition and control") %>% pull(sd) %>% round_x(., 2)`), but reported that such capacities increase dramatically over the target age range (average smoothing term: $F$ = `r round_x(s3_gam_stab["s(target_year):factorCognition and control", "F"], 2)`, $p$ `r ifelse(s3_gam_stab["s(target_year):factorCognition and control", "p-value"] < 0.001, "< 0.001", paste("=", round_x(s3_gam_stab["s(target_year):factorCognition and control", "p"], 3)))`, $edf$ = `r round_x(s3_gam_stab["s(target_year):factorCognition and control", "edf"], 2)`). Again, participants attributed substantially greater capacities to newborns in the domains of _social connection_ (median score for newborns: `r s3_cap_rat_newborns_ss %>% filter(factor == "Social connection") %>% pull(median) %>% round_x(., 2)` out of 100 points, $M$ = `r s3_cap_rat_newborns_ss %>% filter(factor == "Social connection") %>% pull(mean) %>% round_x(., 2)`, $sd$ = `r s3_cap_rat_newborns_ss %>% filter(factor == "Social connection") %>% pull(sd) %>% round_x(., 2)`; increase in log-odds: $\beta$ = `r round_x(s3_gam_ptab["factorsoc_cog", "Estimate"], 2)`, $p$ `r ifelse(s3_gam_ptab["factorsoc_cog", "Pr(>|t|)"] < 0.001, "< 0.001", paste("=", round_x(s3_gam_ptab["factorsoc_cog", "Pr(>|t|)"], 3)))`) and _negative affect_ (median score for newborns: `r s3_cap_rat_newborns_ss %>% filter(factor == "Negative affect") %>% pull(median) %>% round_x(., 2)` out of 100 points, $M$ = `r s3_cap_rat_newborns_ss %>% filter(factor == "Negative affect") %>% pull(mean) %>% round_x(., 2)`, $sd$ = `r s3_cap_rat_newborns_ss %>% filter(factor == "Negative affect") %>% pull(sd) %>% round_x(., 2)`; increase in log-odds: $\beta$ = `r round_x(s3_gam_ptab["factorneg_cog", "Estimate"], 2)`, $p$ `r ifelse(s3_gam_ptab["factorneg_cog", "Pr(>|t|)"] < 0.001, "< 0.001", paste("=", round_x(s3_gam_ptab["factorneg_cog", "Pr(>|t|)"], 3)))`); in Study 3, as in Study 2, participants reported relatively small increases over the target age range in the domain of _negative affect_ (average smoothing term: $F$ = `r round_x(s3_gam_stab["s(target_year):factorNegative affect", "F"], 2)`, $p$ `r ifelse(s3_gam_stab["s(target_year):factorNegative affect", "p-value"] < 0.001, "< 0.001", paste("=", round_x(s3_gam_stab["s(target_year):factorNegative affect", "p"], 3)))`, $edf$ = `r round_x(s3_gam_stab["s(target_year):factorCognition and control", "edf"], 2)`), but in this case they reported quite dramatic increases in the domain of _social connection_ ($F$ = `r round_x(s3_gam_stab["s(target_year):factorSocial connection", "F"], 2)`, $p$ `r ifelse(s3_gam_stab["s(target_year):factorSocial connection", "p-value"] < 0.001, "< 0.001", paste("=",  round_x(s3_gam_stab["s(target_year):factorCognition and control", "p"], 3)))`, $edf$ = `r round_x(s3_gam_stab["s(target_year):factorSocial connection", "edf"], 2)`). Finally, as in Study 2, in the domain of _bodily sensations_ participants attributed nearly maximal capacity to newborns (median score for newborns: `r s3_cap_rat_newborns_ss %>% filter(factor == "Bodily sensations") %>% pull(median) %>% round_x(., 2)` out of 100 points, $M$ = `r s3_cap_rat_newborns_ss %>% filter(factor == "Bodily sensations") %>% pull(mean) %>% round_x(., 2)`, $sd$ = `r s3_cap_rat_newborns_ss %>% filter(factor == "Bodily sensations") %>% pull(sd) %>% round_x(., 2)`; increase in log-odds relative to _cognition and control_: $\beta$ = `r round_x(s3_gam_ptab["factorbod_cog", "Estimate"], 2)`, $p$ `r ifelse(s3_gam_ptab["factorbod_cog", "Pr(>|t|)"] < 0.001, "< 0.001", paste("=", round_x(s3_gam_ptab["factorbod_cog", "Pr(>|t|)"], 3)))`) and a significant but small increase in capacities for _bodily sensations_ over the target age range (average smoothing term: $F$ = `r round_x(s3_gam_stab["s(target_year):factorBodily sensations", "F"], 2)`, $p$ `r ifelse(s3_gam_stab["s(target_year):factorBodily sensations", "p-value"] < 0.001, "< 0.001", paste("=", round_x(s3_gam_stab["s(target_year):factorBodily sensations", "p"], 3)))`, $edf$ = `r round_x(s3_gam_stab["s(target_year):factorBodily sensations", "edf"], 2)`).

>>>>

---
title: "Beliefs about the development of mental life"
author: "[authors removed for anonymous review]"
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
    fig_caption: yes
    toc: false
    number_sections: false
  #   reference_docx: word-styles-ref.docx
  # word_document:
bibliography: "../supplement/references.bib"
csl: "../supplement/apa.csl"
# cls: "../supplement/article.cls"
figsintext: yes
header-includes: 
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{float}
---

@@

Caregiving relationships with infants and children are among the most common, important, and complex human social interactions. Caring for a young child over time requires an adult to continually update their own mental representations of the child’s capacities as the child develops. In a series of large-scale studies, we assessed how US adults conceptualize the development of the human mind over the first five years of life. Exploratory factor analysis identified four categories of mental capacities that anchored participants’ representations of the developing human mind: _bodily sensation_ (e.g., hunger, pain), _negative affect_ (e.g., distress, frustration), _social connection_ (e.g., love, learning from others), and _cognition and control_ (e.g., planning, self-control). Participants perceived that these four aspects of mental life were present to different degrees at birth, followed different developmental trajectories, and were driven by different developmental mechanisms (such as biological "preprogramming," physical maturation, passive observation, and social learning). Adults’ perceptions of children’s mental capacities have important consequences for the well-being of children in their care—particularly in the first few years of life, when children’s communication skills are limited and caregivers must infer children’s thoughts, feelings, and needs. The current studies shed light on this fascinating and understudied aspect of "mind perception," in turn laying the foundation for theory-based interventions to encourage positive and responsive parenting behaviors.

\newpage
Frustration, compassion, worry, humor, imagination, love—such experiences are fundamental parts of human life, but few people would assert that these capacities are fully developed at birth. In any given interaction with a child, adults come face to face with a being whose experience of the world is like their own, but different; whose mind is capable of some of the thoughts and feelings that an adult’s mind is, but perhaps not all of them. When an adult interacts with the same child over time, they witness innumerable changes in that child’s mental life: A sleepy newborn gradually becomes a child who experiences complex emotions and original ideas and develops personal memories and intricate plans. This familiar but remarkable transformation requires adults to make inferences about a given child’s developing mental capacities in order to interpret that child’s behaviors and respond appropriately.

Indeed, caring for a young child involves making countless decisions under uncertainty: Why is my child crying, smiling, shouting, or biting? What do they want, and what do they need? Can they understand what I am saying? Do they love me? As in other domains of reasoning [@gerstenberg.tenenbaum2017; @gopnik.wellman1994; @wellman.gelman1992], people likely draw on folk beliefs and intuitive theories about how the mind develops to predict, explain, and respond to children’s behavior and development, with consequences for the well-being of children in their care.

These folk beliefs and intuitive theories are, to date, poorly understood by social scientists. A handful of recent studies have examined the extent to which cognitive abilities, knowledge, and emotions are thought to be innate vs. learned, with special attention to comparing laypeople’s beliefs about innateness to scientific evidence for or against innateness and universality [@wang.feigenson2019; @berent.etal2019; @berent.etal2020; see @berent2021 for discussion]. These studies offer a fascinating glimpse of potential gaps, misconceptions, and biases in laypeople’s beliefs about the human mind, but they leave open many questions about the nature of the folk beliefs and intuitive theories themselves. What is the cognitive architecture underlying laypeople’s beliefs about the origins of the human mind? How do laypeople "carve up" the various sensations, experiences, and abilities that make up human mental life, and—most crucially from the perspective of caregiving—how do laypeople understand changes in mental life across early development?

In this paper, we examine how adults conceptualize the developing minds of infants and children by drawing on techniques developed in studies of mind perception, in which researchers use large datasets and modern statistical methods to reveal conceptual representations of "the mind," broadly construed. In a groundbreaking study, Gray et al. asked participants to compare the mental capacities of a range of humans, animals, technologies, and other entities, and argued that participants’ judgments revealed a fundamental distinction between "agency" (e.g., capacities for self-control, moral reasoning, memory, and planning) and "experience" (e.g., capacities for hunger, fear, pain, and pleasure) [@gray.etal2007]. More recently, Weisman et al. used a similar empirical approach to argue that people’s conceptions of mental life are anchored by three fundamental components: "body" (physiological sensations such as hunger and pain), "heart" (social-emotional abilities such as embarrassment and pride), and "mind" (perceptual-cognitive capacities such as vision and memory [@weisman.etal2017a; see @malle2019 for a similar three-factor solution]. Such findings have inspired subsequent work on topics as wide-ranging as moral reasoning, dehumanization and stereotyping, beliefs about the afterlife, and human-robot interactions.

However, the burgeoning field of mind perception has largely neglected the question of how the human mind changes over development. In many studies, infants and children have simply not been included as targets for mental capacity attributions; instead, participants have been asked to reason about human adults alongside a range of non-human entities, including other animals, technologies, and supernatural beings. Even studies that have probed participants’ mental capacity attributions to infants and children have generally treated them as beings distinct from adults, rather than as snapshots of a single mind as it emerges and changes over time. For example, Gray et al.’s early work on mind perception suggested that a five-month-old infant is an entity that is generally considered to have experience but not agency—in contrast to, say, a five-year-old child (considered to have experience as well as limited agency), a human adult (considered to have maximal experience and agency), or a robot [considered to have agency but not experience; @gray.etal2007]. Likewise, Weisman et al.’s studies characterized infants and young children—along with, e.g., chimpanzees, elephants, and dolphins—as entities that are perceived to be fully capable of physiological sensations but more limited in their perceptual-cognitive capacities and social-emotional abilities; this is in contrast to human adults, who are perceived to be fully capable across these domains [@weisman.etal2017a]. Although such findings can be reinterpreted as providing preliminary evidence that people consider the minds of infants and children to differ from the minds of adults, they do not address how people conceptualize the development of the mind in early life.

Yet infants and young children are far from unusual social partners. Caregiving relationships are at the core of human existence; for many people, infants and young children are among the most frequent and highly valued minds that we encounter. In the current studies we set out to shed new light on this important and understudied aspect of mind perception: how laypeople reason about the development of human mental life. In three studies, we assessed US adults’ folk beliefs about children’s developing abilities for physiological sensation, perception, cognition, emotion, self-regulation, social interaction, and other aspects of "mental life," broadly construed. We focused on development between birth and age five years—a time of particularly dramatic maturation in perceptual acuity, emotional experience, cognitive ability, executive function, social engagement, and many other domains.

Methods, inclusion/exclusion criteria, and analyses marked as "preregistered" were preregistered on the Open Science Framework (OSF) website: (anonymized links for peer review: Study 1: https://osf.io/e6ajh/?view_only=f2539dbce3664172907aa7a381f62d34; Study 2: https://osf.io/j72dg/?view_only=7e4c01fec32f40a3b13c0ebc2b7ced18; Study 3: https://osf.io/xh8ce/?view_only=56649aff37aa4f1db2dc06b2e69576f4). All studies were approved by [IRB information removed to preserve anonymity]. For extended descriptions of methods and results, see Supplemental Material.

# Study 1

We began with an expansive exploration of US adults’ attributions of mental life to infants and young children of different ages. Drawing inspiration from previous work on mind perception [@gray.etal2007; @weisman.etal2017a; @weisman.etal2017b; @weisman.etal2018] as well as standard assessments of infant temperament and behavior [@putnam.etal2006; @rothbart1978], we included a wide range of 60 capacities in this initial exploration, from basic physiological sensations, to capacities for perception, cognition, and emotion, to abilities for self-regulation and social interaction. We asked participants to assess these capacities at three different ages (at birth, 9 months, and 5 years), with the goal of characterizing the conceptual structure underlying participants’ reasoning about the development of mental life: What do people perceive to be the fundamental components of mental life, as they pertain to a developing human child?

We designed Study 1 to explore the correlational structure of participants’ responses, as a window into the underlying conceptual structure that might support their reasoning about the development of mental life. Our primary analysis was a preregistered exploratory factor analysis (EFA) of participants’ capacity ratings. This analysis allowed us to examine which capacities tend to "hang together" in participants’ assessments of the mental lives of infants and young children. For example, when a participant indicated that newborns are highly capable of feeling scared, what other capacities did they tend to attribute to newborns? Following previous work on mind perception, we argue that the suites of capacities revealed by applying dimensionality reduction (in our case, EFA) to this covariance structure offer a meaningful approximation of the latent conceptual structure underlying participants’ reasoning about the developing human mind.

## Methods
